Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
record inference fix

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semalt}[1]{\llangl #1 \rrangl}

@@

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

@@

\begin{definition}
We write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
The shape preference relation is defined as a transitive reflexive closure of the following rules:

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{float}\,&\sqsupseteq\,\ident{int}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \kvd{nullable}\langl\hat{\sigma}\rangl &\sqsupseteq \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsupseteq 
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsupseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &\sqsupseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsupseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \sigma &\sqsupseteq \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &\sqsupseteq \sigma 
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsupseteq \sigma_i')
\end{align}
\vspace{-1.5em}
\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \geq n)
\end{align}
\vspace{-1.5em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\begin{equation*}
\hspace{4em}
\inference[(record)\;]
  { (\nu_i = \nu'_j) \Leftrightarrow (i = j) \wedge (i \leq k)
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 \!:\! \sigma_1,  \; \ldots \;, \nu_k \!:\! \sigma_k, \; \ldots \;, \nu_n \!:\! \tau_n \} \tsep
    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_k \!:\! \sigma'_k, \; \ldots \;, \nu'_m \!:\! \tau'_m \} \vdash\\
    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_k, 
                            \nu_{k+1} \!:\! \addopt{\sigma_{k+1}}, \ldots, \nu_n \!:\! \addopt{\sigma_n},
                            \nu'_{k+1} \!:\! \addopt{\sigma'_{k+1}}, \ldots, \nu'_m \!:\! \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}
\vspace{-2em}

@@

\caption{Inference judgements that define the common preferred shape relation}
\label{fig:subtyping-cst}
\end{figure*}

@@

\item The record shapes are covariant (\ref{eq:sub-record1}) and preferred record can have 
  additional fields (\ref{eq:sub-record2}). The rule (\ref{eq:sub-record3}) implies that records 
  can omit fields, provided that \kvd{null} is a valid value for the 
  field (i.e. the field is optional).
\end{itemize}

\noindent
The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
important. This allows us to prefer records with optional fields over the \kvd{any} 
shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we find a common shape $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
This is a least upper bound and more usable than the top shape $\kvd{var}$.

@@

Given two shapes, the \emph{common preferred shape} relation finds a least upper bound
(Theorem~\ref{thm:lub}). It prefers records, which is important for usability as discussed
earlier.

\begin{definition}
A \emph{common preferred shape} of shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$, obtained according to the inference rules in 
Figure~\ref{fig:subtyping-cst}.
\end{definition}

\noindent
We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
top shape. When finding a common shape of two records (\emph{record}), we return a record that has 
the union of their fields. The shape of shared fields becomes the common preferred shape of their 
respective shapes; fields that are present in only one record are marked as nullable using 
the function $\addopt{-}$.

@@

\paragraph{Properties.}
The set of shapes does not have a \emph{unique} least upper bound, but it 
has a least upper bound with respect to an equivalence between mutually preferred shapes. 
The common preferred shape relation defines a function (Theorem~\ref{thm:func}) that finds a
least upper bound (Theorem~\ref{thm:lub}).

The equivalence on shapes groups records by their non-nullable fields. Given 
$\sigma_1 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\kvd{nullable}\langl\ident{int}\rangl \}$,
$\sigma_2 = \ident{A}~\{ \ident{a}\!:\!\ident{int} \}$ and
$\sigma_3 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\ident{int} \}$,
it is the case that $\sigma_1 \sqsubseteq \sigma_2$ and $\sigma_1 \sqsupseteq \sigma_2$ but
$\sigma_3 \sqsubseteq \sigma_1$ and not vice versa.

Similarly to labels on the labelled top type (\S\ref{sec:inference-vars}), the nullable fields 
of records serve as annotations that provide access to additional values that may or may not be 
available in the input. In the provided code, these always have to be protected by a runtime check.

\begin{theorem}[Preferred shape function]
\label{thm:func}
For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
\end{theorem}
\begin{proof}
The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
preserved recursively and all rules, symmetry does not enable different derivations, even for
non-symmetric (\emph{nullable}) and (\emph{prim}).
\end{proof}

\begin{theorem}[Least upper bound]
\label{thm:lub}
If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound, i.e. 
$\sigma \sqsupseteq \sigma_1$ and $\sigma \sqsupseteq \sigma_2$ and for all $\sigma'$ such that $\sigma' \sqsupseteq \sigma_1$
and $\sigma' \sqsupseteq \sigma_2$, it holds that $\sigma' \sqsupseteq \sigma$.
\end{theorem}
\begin{proof}
By induction over $\vdash$. The algorithm only infers the top shape \kvd{any} when for non-nullable
shapes of distinct tags and so no better preferred shape exists.
\end{proof}

\noindent
Next, we extend the core model (sufficient for the relativized safety) 
with \emph{labelled top types} discussed earlier.

@@

\subsection{Adding labelled top type}
\label{sec:inference-vars}

When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
However, the type providers need more information to provide typed access to the possible 
alternative shapes of data, such as XML nodes. To support track the possible shapes, we label 
the \kvd{any} shape:
%
\begin{equation*}
\sigma = \ldots \lsep \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl
\end{equation*}
%
The shapes $\sigma_1, \ldots, \sigma_n$ represent statically known shapes that appear in the
sample and that we expose in the provided type. As discussed earlier (\S\ref{sec:providers-xml})
this is important when reading external \emph{open world} data. The labels do not affect the 
preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
seen as the top type, regardless of the labels.

@@

The three rules in Figure~\ref{fig:subtyping-cst-var} replace the earlier (\emph{top}). When 
combining a top with another shape (\emph{top-1}), the labelled top shape may or may not already 
contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
a new case is added. When combining two top types (\emph{top-2}), we group the labels that have 
shared tags. Finally, (\emph{top-3}) covers the case when we are combining two distinct non-top 
shapes. As top shapes implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ 
to make nullable shapes non-nullable (when possible) to simplify the label.

@@

Stating properties of the labels requires a finer relation than \emph{preferred shape}. In particular, 
it can only be done with respect to the \emph{sample}. We leave the details to future work, but we
note that the algorithm infers the best labels in the sense that there are labels that enable 
typed access to every possible value in the sample, but not more. The same is the case for nullable
fields of records.

\subsection{Inferring shapes from values}
\label{sec:formal-inferval}

The common preferred shape relation is at the core of the shape inference. What remains to be 
specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
%
\begin{equation*}
\begin{array}{lcl}
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The first few cases represent primitive values ($i$ for integers, $f$ for floating
point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

The following defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
data $d_1, \ldots, d_n$ into a shape $\sigma$:
%
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i
 \\[0.5em]
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n} \}
\end{array}
\end{equation*}
%
We overload the notation and write $\semalt{s}$ when inferring shape from a single value. Primitive 
values are mapped to their corresponding shapes. For records, we return infer field shapes
from the individual values.

When infering a shape from multiple samples, we use the common preferred shape relation to find a 
common shape for all values (starting with $\bot$). This operation is used at the top-level 
(when calling a type provider with multiple samples) and also when inferring the shape of collection
elements.

@@

This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
which is an object-oriented subset of F\# based on SML \cite{sml} and Featherweight Java \cite{fwjava}. It 
includes operations for working with data that model the F\# Data runtime. Finally, we describe how 
type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).

Type providers map the dirty world of data into a nice world of F\#. To model this, the Foo calculus
does not have \kvd{null} values and data values $d$ are never directly exposed.

@@

The type \ident{Data} is the type of all structural data $d$. A class definition $L$ consists of 
a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
$e$ include class construction, member access and the usual functional constructs (functions,
lists, options) and conditionals. The $op$ constructs are discussed next.

@@

The Foo programs can only work with raw data using certain primitive operations. Those are modelled
by the $op$ primitives. In F\# Data, those are internal and users never access them directly.

>>>>

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semalt}[1]{S(#1)}

@@

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \addopt{\rho_i}  \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

During inference we use row-variables $\rho_i$ to represent the flexibility
implied by our interpretation of records in samples. 
When a record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
it may be combined in a list with other samples that contain more fields. The
overall shape inferred must account for the potential that any other fields are not
present in the first sample. The solutions of $\rho_i$ variables is 
discussed in \S\ref{sec:inference}. We assume substitution and row-unification is
defined for these in the usual way , except that as substitutions are made
the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
in the vector. After inference any remaining $\rho_i$ variables are ignored (assumed empty).

@@

\begin{definition}
For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
The shape preference relation is defined as a transitive reflexive closure of the following rules:

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{float}\,&\sqsupseteq\,\ident{int}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \kvd{nullable}\langl\hat{\sigma}\rangl &\sqsupseteq \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsupseteq 
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsupseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &\sqsupseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsupseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \sigma &\sqsupseteq \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \kvd{any} &\sqsupseteq \sigma 
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsupseteq \sigma_i')
\end{align}

\vspace{-1.5em}

\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \geq n)
\end{align}

\end{definition}

@@

\begin{equation*}
\hspace{4em}
\inference[(record)\;]
  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
    \end{array} }
\end{equation*}
\vspace{-2em}

@@

\caption{The common preferred shape relation}
\label{fig:subtyping-cst}
\end{figure*}

@@

\item The record shapes are covariant (\ref{eq:sub-record1}) and preferred record can have 
  additional fields (\ref{eq:sub-record2}). 
\end{itemize}

\noindent

@@

Given two ground shapes, the \emph{common preferred shape} relation finds a least upper bound
(Theorem~\ref{thm:lub}). It prefers records, which is important for usability as discussed
earlier.

\begin{definition}
A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$, by the rules in 
Figure~\ref{fig:subtyping-cst}.
\end{definition}

\noindent
We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
top shape. When finding a common shape of two records (\emph{record}), the shape of fields becomes the common preferred shape of their 
respective shapes. If the field labels of two record types do not match,
we can assume a common assignment of fields has been determined
through shape inference (\S\ref{sec:formal-inferval}).

@@

\paragraph{Properties.}
A set of ground shapes has a \emph{unique} least upper bound by the preferred shape ordering.
For ground shapes, the common preferred shape relation defines a function (Lemma~\ref{thm:func}) that finds 
the least upper bound (Lemma~\ref{thm:lub}).

\begin{lemma}[Preferred shape function]
\label{thm:func}
For all ground $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
\end{lemma}
\begin{proof}
The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
preserved recursively and all rules, symmetry does not enable different derivations, even for
non-symmetric (\emph{nullable}) and (\emph{prim}).
\end{proof}

\begin{lemma}[Least upper bound]
\label{thm:lub}
For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
\end{lemma}
\begin{proof}
By induction over $\vdash$. The algorithm only infers the top shape \kvd{any} when for non-nullable
shapes of distinct tags and so no better preferred shape exists.
\end{proof}

\subsection{Inferring shapes from samples}
\label{sec:formal-inferval}

What remains to be 
specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
%
\begin{equation*}
\begin{array}{lcl}
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The first few cases of represent primitive values ($i$ for integers, $f$ for floating
point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
data $d_1, \ldots, d_n$ into a shape $\sigma$. We assume each record in each $d_i$ is marked with a fresh row inference
variable $\rho_i$.  We then choose a minimal solution $\theta$ for these row variables
in order to equate the record 
field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
This minimization is implemented via row variable unification, we omit the details here.

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.1em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference}
\label{fig:shape inference}
\end{figure}

%
Primitive  values are mapped to their corresponding shapes. 
When infering a shape from multiple samples, we use the common preferred shape relation to find a 
common shape for all values (starting with $\bot$). This operation is used at the top-level 
(when calling a type provider with multiple samples) and also when inferring the shape of collection
elements.

@@

\subsection{Adding labelled top shapes}
\label{sec:inference-vars}

Next, we indicate how the core model (sufficient for the relativized safety) 
can be extended with \emph{labelled top shapes} discussed earlier.

When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
However, the type providers need more information to provide typed access to the possible 
alternative shapes of data, such as XML nodes. To support track the possible shapes, we label 
the \kvd{any} shape:
%
\begin{equation*}
\sigma = \ldots \lsep \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl
\end{equation*}
%
The shapes $\sigma_1, \ldots, \sigma_n$ represent statically known shapes that appear in the
sample and that we expose in the provided type. As discussed earlier (\S\ref{sec:providers-xml})
this is important when reading external \emph{open world} data. The labels do not affect the 
preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
seen as the top shape, regardless of the labels.

@@

The three rules in Figure~\ref{fig:subtyping-cst-var} replace the earlier (\emph{top}). When 
combining a top with another shape (\emph{top-1}), the labelled top shape may or may not already 
contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
a new case is added. When combining two top shapes (\emph{top-2}), we group the labels that have 
shared tags. Finally, (\emph{top-3}) covers the case when we are combining two distinct non-top 
shapes. As top shapes implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ 
to make nullable shapes non-nullable (when possible) to simplify the label.

@@

Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
We leave the details to future work, but we
note that the algorithm infers the best labels in the sense that there are labels that enable 
typed access to every possible value in the sample, but not more. The same is the case for nullable
fields of records.

@@

This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
which is a subset of F\# using object-property notation with
added operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).

Type providers map the ``dirty'' world of weakly typed structured data into a nice world of strong types. To model this, the Foo calculus
does not have \kvd{null} values and data values $d$ are never directly exposed.  
Furthermore Foo is simply typed: despite using class types and object notation 
for properties, it has no subtyping.

@@

The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
$e$ include class construction, member access and the usual functional constructs (functions,
lists, options) and conditionals. The $op$ constructs are discussed next.

@@

The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
